---
title: "PM566 Assignment 4"
author: "Echo Tang"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## HPC

### Problem 1 - Making Code Nicer
```{r}
fun1 <- function(mat) {
  n <- nrow(mat)
  ans <- double(n) 
  for (i in 1:n) {
    ans[i] <- sum(mat[i, ])
  }
  ans
}

fun1alt <- function(mat) {
  ans = rowSums(mat)
  ans
}


fun2 <- function(mat) {
  n <- nrow(mat)
  k <- ncol(mat)
  ans <- mat
  for (i in 1:n) {
    for (j in 2:k) {
      ans[i,j] <- mat[i, j] + ans[i, j - 1]
    }
  }
  ans
}



fun2alt <- function(mat) {
  ans = matrix(nrow = nrow(mat), ncol = ncol(mat))
  for(i in 1:nrow(mat)){
    temp = mat[i, ]
    ans[i, ] = cumsum(temp)
  }
  ans
}

# Use the data with this code
set.seed(2315)
dat <- matrix(rnorm(200 * 100), nrow = 200)

# Test for the first
microbenchmark::microbenchmark(
  fun1(dat),
  fun1alt(dat), unit =  "ms", check = "equivalent"
)


# Test for the second
microbenchmark::microbenchmark(
  fun2(dat),
  fun2alt(dat), unit = "ms", check = "equivalent"
)
```

### Problem 2 - Make things run faster with parallel computing
```{r}
sim_pi <- function(n = 1000, i = NULL) {
  p <- matrix(runif(n*2), ncol = 2)
  mean(rowSums(p^2) < 1) * 4
}

set.seed(1231)
system.time({
  ans <- unlist(lapply(1:4000, sim_pi, n = 10000))
  print(mean(ans))
})

system.time({
  # YOUR CODE HERE
  cl <- parallel::makePSOCKcluster(4)
  parallel::clusterSetRNGStream(cl, 1231)
  parallel::clusterExport(cl, c("sim_pi"), envir=environment())
  ans <- unlist(parallel::parLapply(cl, 1:4000, sim_pi, n = 10000))
  print(mean(ans))
  # YOUR CODE HERE
  
})
```

## SQL
```{r}
library(RSQLite)
library(DBI)

con <- dbConnect(SQLite(), ":memory:")

film <- read.csv("https://raw.githubusercontent.com/ivanceras/sakila/master/csv-sakila-db/film.csv")
film_category <- read.csv("https://raw.githubusercontent.com/ivanceras/sakila/master/csv-sakila-db/film_category.csv")
category <- read.csv("https://raw.githubusercontent.com/ivanceras/sakila/master/csv-sakila-db/category.csv")

dbWriteTable(con, "film", film)
dbWriteTable(con, "film_category", film_category)
dbWriteTable(con, "category", category)

```

### Question 1
```{sql, connection = con}
SELECT rating, COUNT(rating) as count
FROM film
GROUP BY rating
ORDER by count

```


### Question 2
```{sql, connection = con}
SELECT rating, AVG(replacement_cost) as avg_rep, AVG(rental_rate) as avg_ren
FROM film
GROUP BY rating

```

The average replacement cost and rental rate for G rated movies are 20.12 and 2.91 respectively. The average replacement cost and rental rate for NC-17 movies are 20.13 and 2.97 respectively. The average replacement cost and rental rate for PG rated movies are 18.96 and 3.05 respectively. The average replacement cost and rental rate for PG-13 rated movies are 20.40 and 3.03 respectively. The average replacement cost and rental rate for R rated movies are 20.23 and 2.94 respectively.

### Question 3
```{sql, connection = con}
SELECT category_id, COUNT(b.film_id) as count
FROM film AS a
  INNER JOIN film_category AS b
  ON a.film_id = b.film_id
GROUP BY category_id

```

### Question 4
```{sql, connection = con}
SELECT a.category_id, COUNT(a.film_id) as count, name
FROM film_category AS a
  INNER JOIN category AS b
  ON a.category_id = b.category_id
GROUP BY name
ORDER BY count DESC
```

The Sports category has the most films with 74 films, followed by Foreign, which has 73 films. 
